Columnar VCF for Biobank scale datasets using Zarr

### 2024-04-08 paper.tex

\subsection{Storing the genotype matrix}
Although VCF is the standard format for exchanging genetic variation
data, its limitations both in terms of compression ratios
and query/compute performance are well known, and many methods 
have been suggested to improve on these properties.[REVIEW METHODS]
High level discussion of
strategies used to store genotype matrix and accompanying metadata.

\begin{figure}
    \begin{tabular}{c}
\resizebox{7cm}{!}{\input{diagrams/data-model.tex}} \\
\includegraphics[width=7cm]{figures/data-scaling}
\end{tabular}
\caption{Chunked compressed storage of the genotype matrix using Zarr.
(A) Schematic of the chunked array storage used in sgkit
(B) Compression performance compared with BCF, savvy and genozip
based on simulated data (see text for details).
\label{fig-data-storage}}
\end{figure}

Emphasis on access by variant, not sample. Introduce Zarr, and
refer to Fig~\ref{fig-data-storage}A to explain columnar storage,
and Fig~\ref{fig-data-storage}B to show that is competitive
with highly specialised approaches.
Note that the FC simulated dataset is probably a best-case
scenario for methods specialised to patterns in the genotype matrix.
Discuss storing additional data, and advantages of the
columnar approach (refs to genomics tools that do the same).

\subsection{Computing on the genotype matrix}
Compressing data isn't ultimately the point - we want to use it.
We discuss the problem of running computations on the entire
genotype matrix, and how different formats limit what can be done.
We illustrate the discussion by the afdist operation from
the bcftools plugins package. This is perhaps a contrived example, but
we chose it because it requires examining the entire genotype matrix
and there is an efficient implementation available in bcftools.
It is also a good example because the standard approach of
divvying up the genome into chunks via HPC schedulers or workflow
managers is quite tricky.
We discuss the software development aspect, and how we developed
the afdist-savvy program. We discuss the inherent limitations
of the decode-to-vcf strategy, and how this inevitably
leads to bottlenecks.

@@

Blosc is a ``meta-compressor'' because it provides different 
access to several different compression codecs. The 
Zstandard compressor~\citep{collet2021rfc} is of particular 
interest here as it achieves very high compression ratios
with fast decompression speeds [see fig]. 
% This may seem a bit off topic, but want to reassure reader that 
% this isn't some niche technology
Zstandard is widely
used and is being incorporated, e.g., as a standard compression
mechanism for HTTP traffic and to decompress Linux kernel images
during boot.
% LZ4 is also notable - any citations for this?

@@

As has been noted many times~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},

The shortcomings of VCF (and to a lesser degree, BCF) as a storage
format and the basis of efficient computation have long been
recognised
and many alternatives
have been proposed.

Purely archival formats try to get the maximum possible compression levels,
without much attention to query performance. Typically these use
highly specialised approaches to maximise compression performance
on genotype data.

These include:
TGC~\citep{deorowicz2013genome}
GTShark~\citep{deorowicz2019gtshark},
VCFShark~\citep{deorowicz2021vcfshark}.

SpVCF~\citep{lin2020sparse} is an extension of the VCF text
format that maximises compatibility, while gaining substantial
improvements in file sizes.

Many tools focus on balancing compression performance with
being able to run queries, typically outputting a VCF.
e.g.
BGT~\citep{li2016bgt},
GQT~\citep{layer2016efficient},
SAVVY~\citep{lefaive2021sparse},
XSI~\citep{wertenbroek2022xsi},
GTC~\citep{danek2018gtc},
GTRAC~\citep{tatwawadi2016gtrac}
genozip~\citep{lan2020genozip,lan2021genozip}
GBC~\citep{zhang2023gbc}.

@@

Many recent tools are based on the Positional Burrows--Wheeler
Transform (PBWT)~\citep{durbin2014efficient}, which takes advantage
of the underlying haplotype structure of genomic data to provide
concise storage, and efficient haplotype matching.

>>>>

\subsection{Storing genetic variation data}
Although VCF is the standard format for exchanging genetic variation
data, its limitations both in terms of compression 
and query/compute performance are well 
known~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
and many methods 
have been suggested to improve on these properties.
Most approaches balance compression with
performance on particular types of queries, 
typically using a command line interface
and outputting VCF text~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
Several specialised algorithms for compressing 
the genotype matrix (i.e., just the genotype calls without additional
VCF information) have been proposed
\citep{deorowicz2013genome, %TGC
deorowicz2019gtshark, %GTShark
deorowicz2021vcfshark}, % VCFShark
most notably the highly-influential Positional
Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.

VCF is a text-based format in which each line
describes observations
and metadata for a single variant (usually a position on the genome
where variation occurs in the samples)~\citep{danecek2011variant}. 
Blocks of VCF text are compressed using \texttt{bgzip},
which allows subsets of the file to be retrieved using secondary
indexes stored alongside the original file~\citep{li2011tabix}.
VCF is therefore a \emph{row-wise} format, i.e., where all of the 
data for a record is stored continguously. BCF is similarly 
row-wise, as are the majority of proposed alternative storage formats.
Row-wise storage makes retrieving all information
for a given record straightforward and efficient, and 
allows data to be easily streamed through Unix pipelines (see Discussion).

Row-wise storage works well when records are relatively small, 
or when we typically want to analyse each record in its entirity.
When we want to analyse only a subset of a record 
row-wise storage can inefficient because we will usually need to
retrieve more information than required from storage. In the case 
of VCF (and BCF) where records are not of a fixed size and 
are almost always compressed in blocks, accessing any information
for all rows means retrieving and decompressing \emph{all} 
information from all rows.
The alternative to row-wise storage is \emph{columnar} storage:
instead of grouping toether all the fields for a record,
we group together all the records for a given field.
To our knowledge, the only existing method using columnar 
storage for genetic variation data is 
SeqArray~\citep{zheng2017seqarray} built on 
the GDS container format~\citep{zheng2012high}.

\begin{figure}
    \begin{tabular}{c}
\resizebox{7cm}{!}{\input{diagrams/data-model.tex}} \\
\includegraphics[width=7cm]{figures/data-scaling}
\end{tabular}
\caption{Chunked compressed storage of the genotype matrix using Zarr.
(A) Schematic of data stored under the \texttt{vcfzarr} specification.
(B) Compression performance compared with BCF, savvy and genozip
based on simulated data (see text for details).
\label{fig-data-storage}}
\end{figure}

We propose the \texttt{vcfzarr} specification which maps the 
VCF data model into a columnar layout using the
widely-used Zarr format (Fig~\ref{fig-data-storage}A). 
Briefly, Zarr is a format 
% FIXME two different uses of $n$ here in a paragraph
for storing $n$-dimensional arrays of a fixed datatype, stored as 
compressed chunks. In the \texttt{vcfzarr} specification, 
each field in a VCF is mapped to an array, which is compressed 
and stored individually allowing for efficient retrieval. 
In particular, call-level data is stored as $m \times n$ arrays
(for $m$ sites and $n$ samples), allowing for efficient 
retrieval of subsets of those fields along both the 
variants and samples axis.
Note that this approach to storing higher dimensional 
sets Zarr apart from more general columnar formats 
such as Parquet [TODO more details and citations?].
See the Methods for more detail on Zarr and the \texttt{vcfzarr}
specification.

Remarkably, the simple approach of compressing
two dimensional chunks with the Zstandard 
compressor~\citep{collet2021rfc} and the bit-shuffle
filter from Blosc~\cite{alted2010modern} produces 
compression levels competitive with specialised methods
on the genotype matrix. In Fig~\ref{fig-data-storage}B
we show compression performance on up to a million
samples subset a large and highly realistic 
simulation of French-Canadians~\cite{anderson2023on}
(see Methods for details).
Note that while simulations cannot capture 
all the subtleties of real data, the allele frequency
and population structure patterns in this dataset 
have been shown to closely follow 
observations~\cite{anderson2023on} and so provides 
a reasonable data point when comparing such methods.
% Could note here that this might represent a best 
% case scenario for specialised methods, but maybe 
% not worth bothering with.

\subsection{Computing on the genotype matrix}
Compressing data isn't ultimately the point - we want to use it.
We discuss the problem of running computations on the entire
genotype matrix, and how different formats limit what can be done.

[Review computational accessibility of the available formats.
We have those that only produce VCF in response to limited 
query languages. Those with APIs.]

We illustrate the discussion by the afdist operation from
the bcftools plugins package. This is perhaps a contrived example, but
we chose it because it requires examining the entire genotype matrix
and there is an efficient implementation available in bcftools.
It is also a good example because the standard approach of
divvying up the genome into chunks via HPC schedulers or workflow
managers is quite tricky.
We discuss the software development aspect, and how we developed
the afdist-savvy program. We discuss the inherent limitations

of the decode-to-vcf strategy, and how this inevitably
leads to bottlenecks.

@@

Blosc is a ``meta-compressor'' because it provides different 
access to several different compression codecs. The 
Zstandard compressor~\citep{collet2021rfc} is of particular 
interest here as it achieves very high compression ratios
with fast decompression speeds [see fig]. 
% This may seem a bit off topic, but want to reassure reader that 
% this isn't some niche technology
Zstandard is also used in several recent VCF compression 
methods~\citep[e.g.][]{lefaive2021sparse,wertenbroek2022xsi}.
% LZ4 is also notable - any citations for this?

@@

@@
